\section{Literature review}\label{Literature review}

In the last two decades there has been a significant increase in the amount of interest in the potential applications of Artificial  Intelligence (AI) methods in various domains of science. This has also been the case for finance.

\subsection{Artificial Neural Network (ANN) forecasting}\label{Artificial Neural Networks (ANN) forecasting}

\begin{itemize}
    \item In \cite{ann-2} Adewole et al. propose an Artificial Neural Network Foreign Exchange Rate Forecasting model for FOREX rate forecasting that corrects the handling of the uncertainty and instability in FOREX data. When compared against the Hidden Markov foreign exchange rate forecasting model (HFERFM) which showed an accuracy of 69.9\% their model showed an accuracy of 81.2\%.
    \item In \cite{performance} S. Galeshchuk shows that the short-term one step ahead prediction of panel exchange rate data using Neural Networks (NN) provides good accuracy and can be used in practical systems.
    \item In \cite{ann-1} Kamruzzaman and Sarker developed and investigated three Artificial Neural Network (ANN) based forecasting models using Standard Backpropagation (SBP), Scaled Conjugate Gradient (SCG) and Backpropagation with Baysian Regularization (BPR) for Australian Foreign Exchange to predict six different currencies against the Australian dollar. All of their models outperformed the ARIMA model. It was found that the SCG based model performed best when measured with the two most commonly used metrics and showed competitive results when compared with the BPR based model using the remaining three metrics. Their experimental results demonstrate that ANN based models can closely forecast the forex market \cite{ann-1}.
    \item In \cite{gbp-usd-forecasting} Gioqinang and Hu examine the effects of the number of input and hidden nodes as well as the size of the training sample on the in-sample and out-of-sample performance of Neural Networks (NN). It is found that neural networks outperform linear models, particularly when the forecast horizon is short. In addition, the number of input nodes has a greater impact on performance than the number of hidden nodes, while a larger number of observations do reduce forecast errors.
\end{itemize}

\subsection{Random Forest (RF) forecasting}\label{Random Forest (RF) forecasting}

\begin{itemize}
    \item In \cite{prediction} Kaidem et al. propose a novel way to minimize the risk of investment in stock market by predicting the returns of a stock using Random Forest (RF) regression. Their model is shown to outperform existing estimators found in the literature.
    \item In \cite{gold} Krishna et al. utilize a Random Forest (RF) regression model, to predict future trends in gold prices. They use historical data, market indicators, and economic variables for near real-time prediction. This enhances the traditional methods of financial analysis by addressing typical challenges such as market volatility, external economic shocks, and the unpredictability of commodity prices.
    \item In \cite{wavelet} Xu et al. propose a novel approach that combines a Random Forest (RF) with the signal processing method of the wavelet transform to model the prediction of currency crises. The model is shown to achieve a high level of predictive accuracy.
    \item In \cite{bitcoin} Basher and Sadorsky find that random forests predict Bitcoin and gold price directions with a higher degree of accuracy than logit models. Prediction accuracy for bagging and random forests is between 75\% and 80\% for a five-day prediction. For 10-day to 20-day forecasts bagging and random forests record accuracies greater than 85\%. Second, technical indicators are the most important features for predicting Bitcoin and gold price direction, suggesting some degree of market inefficiency. Third, oil price volatility is important for predicting Bitcoin and gold prices indicating that Bitcoin is a substitute for gold in diversifying this type of volatility. By comparison, gold prices are more influenced by inflation than Bitcoin prices, indicating that gold can be used as a hedge or diversification asset against inflation.
\end{itemize}

\subsection{Support Vector Machine (SVM) forecasting}\label{Support Vector Machine (SVM) forecasting}

\begin{itemize}
    \item In \cite{svm-2} Alamili identified the best performing SVM in terms of model structure and parameters for the prediction of the EUR/USD exchange rate. The performance of the SVM model was compared to that of an ANN.
    \item In \cite{svm-1} Kamruzzaman et al. investigate the effect of different kernel functions, namely, linear, polynomial, radial basis and spline on the prediction error measured by several widely used performance metrics. The effect of regularization parameter is also studied. In reducing the total prediction error the polynomial kernel produced the best result while in predicting the trend the radial basis and polynomial kernels produced equally good results. In general the radial basis and polynomial kernel appeared to be the better choice for FOREX market forecasting.
    \item In \cite{svr} Yaohao and Albuquerque verify the predictive capacity of nine SVR models for predicting the spot nominal exchange rate of 10 pairs of currencies: US dollar, euro, British pound sterling, Chinese renminbi and Japanese yen, combined two by two based on a fundamentalist model composed of 13 explanatory variables. The results showed that the majority of SVR models achieved better out-of-sample performance than the Random Walk, but overall failed to achieve statistical significance of predictive superiority.
\end{itemize}

\subsection{Hybrid models}\label{Hybrid models}

\begin{itemize}
    \item In \cite{hybrid-model} Ince and Trafalis propose a two stage forecasting model which incorporates parametric techniques such as autoregressive integrated moving average (ARIMA), vector autoregressive (VAR) and co-integration techniques, and nonparametric techniques such as support vector regression (SVR) and artificial neural networks (ANN). Comparison of the models showed that input selection is very important. Furthermore, it was found that the SVR technique outperforms the ANN for two input selection methods.
    \item In \cite{cgp-rnn} Rehman et al. propose a novel approach to foreign currency exchange rates forecasting based on Recurrent Neuro-Evolution. Recurrent Cartesian Genetic Programming evolved Artificial Neural Network (RCGPANN) is demonstrated to produce a computationally efficient and accurate model for FOREX prediction with an accuracy reaching 98.872\% for a period of 1000 days. The reason for the efficient performance of the system is the ability of the system to select the best possible feature, network architecture and connectivity pattern for prediction and to decide whether to use a recurrent connection or a feed-forward connection.
    \item In \cite{neuro-fuzzy} Yong et al. proposed a new approach called Gaussian Mixture Model Initialized Neuro-Fuzzy (GMMINF) for foreign currency exchange rate prediction. The experimental forecast results for the model for the Close and Low prices for the main three currencies benchmarked in the worst of cases reach a score of 99.5398\% accuracy.
    \item In \cite{rf-svr} Jazayeri uses a stacked ensemble of Random Forest and Support Vector Regression (SVR) to predict future exchange rates. The model is compared against a single Random Forest (RF), a single SVR, and an average ensemble of RF and SVR models. The research method is further fortified by the use of k-fold cross-validation and ANOVA tests. The findings of the study reveal significant predictive accuracy of the stacked ensemble model, emphasizing the intricate interconnections of currency exchange rates.
    \item In \cite{genetic} Pacelli et al. conduct three days ahead forecasting of the EUR/USD exchange rate using ANN designed and tested by means of the specific genetic algorithm multi-objective Pareto-Based and conclude that the model can accurately carry out ex-post predictions.
    \item In \cite{robust-regression} Preminger and Franck conduct a forecast accuracy and sign predictability measure comparison for robust linear autoregression (RAR), robust neural networks (RNN), random walk (RW) and linear autoregression (AR). They find that robust models tend to improve the forecasting accuracy of the AR and of the NN at all time horizons, and even of the RW for forecasts carried out at the 1-month horizon. Robust models are also shown to have significant market timing ability at all forecast horizons.
\end{itemize}

\subsection{Model comparison}\label{Model comparison}

\begin{itemize}
    \item In \cite{nn-vs-chaotic} Lisi and Schiavo conduct a comparative evaluation of chaotic models and neural networks for forecasting currency exchange rates. According to the mean square error criterion, neural networks perform slightly better than chaotic models and, in turn, both methods substantially outperform the random walk model.
    \item In \cite{linear-vs-nn} Binner et al. conduct a comparison of the relative Euro inflation forecasting performance of linear ARIMA and VAR models and the nonlinear NN. In conclusion they find that Neural Networks (NN) have the potential to have better predictive ability than linear models however lack suitable methods for the selection of optimal parameters.
    \item In \cite{techniques} Mohammadi conducts a comparison of four machine learning models; Backpropagation, Radial Basis Function (RBF), Long Short-Term Memory, and Support Vector Regression (SVR). He finds that SVR outperforms the other three models with the Backpropagation neural network being the least performant.
    \item In \cite{rf-svm} Santuci et al. compare the FOREX forecasting performance of the Random Forest and SVM against classic technical indicators, like moving averages. It is found that the proposed algorithms outperform the traditional technical indicators.
    \item In \cite{lstm-mlp-rf} Khan analyses the forecasting of exchange rates relative to foreign currencies using Long Short-Term Memory (LSTM), MultiLayer Perceptron (MLP), and Random Forest for USD/PKR and PKR/USD, MYR/PKR exchange rates. The study shows that LSTM and Random Forest models show promising results in predicting future trends and that both models are incredibly feasible.  Although LSTM is an efficient method for processing sequential data, it has shortcomings in dealing with short-term fluctuations and possible errors. The main disadvantage of MLP models is that despite their benefit in easily fitting the training data, they are not as immune to noise and variations.
    \item In \cite{review} Islam et al. present a comprehensive review of FOREX rate forecasting research from 2017 to 2019. From the reviewed literature, it is clear that neural network-based approaches, optimization techniques, and pattern-based approaches were the most liked techniques among the researchers at the aforementioned time.
\end{itemize}

