\section{Model training and evaluation}\label{Model training and evaluation}

This section presents an overview of the primary guideing principals used during the stages of model training and evaluation.

\subsection{Bias-Variance dilemma}\label{Bias-Variance dilemma}

The bias-variance dilemma is a fundamental problem in machine learning and involves the issues of model optimization and generalization. In the case of optimization (reducing the model error) a model eventually starts to overfit the training data, this results in high error of the models predictions on the test/validation datasets (high variance). The opposite is aiming to increase model generalization (reduction of model complexity) however this in turn causes underfitting or high error of the models predictions on the training dataset (high bias). This counterposed behaviour of both bias and variance enables the easy assessment of model suitability for use in the prediction of unseen data \cite{ibm-ensemble,ibm-regularization,svm-2}. 

\subsection{Performance metrics}\label{Performance metrics}

The accuracy of the model forecasts was evaluated on the basis of commonly used measures including:
\begin{itemize}
    \item \Gls{mape} - calculates the average of the absolute percentage errors, where the percentage error is the difference between the actual and predicted values, divided by the actual value. MAPE provides a percentage-based error metric that is useful for interpreting the overall accuracy of a model, especially when the scale of the target variable varies \cite{scikit-mape}.
    \item \Gls{ds} - is a metric that measures the ability of a forecasting model to correctly predict the direction of change in the target variable. It calculates the percentage of times the model correctly predicts whether the value will increase or decrease compared to the previous period \cite{svm-2}.
    \item \Gls{oob} error - is a method for evaluating the performance of ensemble methods where during training on a random subset of the training data the generated model is evaluated on the data that was not part of its training dataset. It provides an unbiased estimate of the model's generalization performance without the need for a separate validation set \cite{scikit-oob}.
\end{itemize}

% Juxtaposed
% 
% In the case of generalization (reduction of model complexity) a model starts to underfit the training data, this results in high error of the models predictions on the training dataset (high bias).
% 
% The opposite is aiming to increase model optimization (reducing the model error) however this in turn pushed to far causes overfitting or high error of the models predictions on the test/validation datasets (high variance).

\subsection{Methodology}\label{Methodology}

For the purpose of this paper only one of the currency exchange rates was chosen as a target for forecasting alongside five currency exchange rates which were selected as the feature set. The chosen currencies were: \gls{nzd}, \gls{sek}, \gls{cad}, \gls{dkk} and \gls{pln} as predictors and \gls{usd} as the target variable.

\subsubsection{Sample window size selection}\label{Sample window size selection}

In the first step the default parameters of each of the models were used. The only exception to this was the \gls{mlp} where the default activation function \gls{relu} was switched out for \gls{tanh} due to the normalization range of the data. Each models accuracy was evaluated on both in-sample (training set) and out-of-sample (test set) data.

\begin{figure}[ht!]
  \centering
  \begin{tikzpicture}
    \begin{groupplot}[
      group style={
        group size=1 by 2, % 1 column, 2 rows
        vertical sep=2cm,  % space between plots
      },
      width=\linewidth,
      height=8cm,
      grid=major,
      grid style={dashed,gray!30},
      xlabel=Sample window,
      x unit=\si{Days},
      x tick label style={rotate=90,anchor=east},
      legend style={at={(0.5,-0.2)}, anchor=north, legend columns=-1},
      legend to name=sharedlegend
    ]

    \nextgroupplot[ylabel=\gls{mape}, y unit=\si{\%}]
    \addplot table[x=Offset,y=MAPE,col sep=comma] {data/usd/nn/test_mape.csv};
    \addplot table[x=Offset,y=MAPE,col sep=comma] {data/usd/nn/train_mape.csv};

    \nextgroupplot[ylabel=\gls{ds},  y unit=\si{\%}]
    \addplot table[x=Offset,y=Hitrate,col sep=comma] {data/usd/nn/test_hitrate.csv};
    \addplot table[x=Offset,y=Hitrate,col sep=comma] {data/usd/nn/train_hitrate.csv};

    \addlegendentry{Test}
    \addlegendentry{Train}

    \end{groupplot}
  \end{tikzpicture}

  \vspace{1em}
  %\ref{sharedlegend}
  \pgfplotslegendfromname{sharedlegend}
    \caption{\gls{usd} \gls{mlp}: \gls{mape} and \gls{ds} for different sample windows.}
\end{figure}

\begin{figure}[ht!]
  \centering
  \begin{tikzpicture}
    \begin{groupplot}[
      group style={
        group size=1 by 3, % 1 column, 3 rows
        vertical sep=2cm,  % space between plots
      },
      width=\linewidth,
      height=6cm,
      grid=major,
      grid style={dashed,gray!30},
      xlabel=Sample window,
      x unit=\si{Days},
      x tick label style={rotate=90,anchor=east},
      legend style={at={(0.5,-0.2)}, anchor=north, legend columns=-1},
      legend to name=sharedlegend
    ]

    \nextgroupplot[ylabel=\gls{mape}, y unit=\si{\%}]
    \addplot table[x=Offset,y=MAPE,col sep=comma] {data/usd/rf/test_mape.csv};
    \addplot table[x=Offset,y=MAPE,col sep=comma] {data/usd/rf/train_mape.csv};

    \nextgroupplot[ylabel=\gls{ds},  y unit=\si{\%}]
    \addplot table[x=Offset,y=Hitrate,col sep=comma] {data/usd/rf/test_hitrate.csv};
    \addplot table[x=Offset,y=Hitrate,col sep=comma] {data/usd/rf/train_hitrate.csv};

    \nextgroupplot[ylabel=\gls{oob} Error]
    \addplot table[x=Offset,y=OOB Error,col sep=comma] {data/usd/rf/oob_error.csv};

    \addlegendentry{Test}
    \addlegendentry{Train}

    \end{groupplot}
  \end{tikzpicture}

  \vspace{1em}
  %\ref{sharedlegend}
  \pgfplotslegendfromname{sharedlegend}
  \caption{\gls{usd} \gls{rf}: \gls{mape} and \gls{ds} for different sample windows.}
\end{figure}

\begin{figure}[ht!]
  \centering
  \begin{tikzpicture}
    \begin{groupplot}[
      group style={
        group size=1 by 2, % 1 column, 2 rows
        vertical sep=2cm,  % space between plots
      },
      width=\linewidth,
      height=8cm,
      grid=major,
      grid style={dashed,gray!30},
        xlabel=Sample window,
      x unit=\si{Days},
      x tick label style={rotate=90,anchor=east},
      legend style={at={(0.5,-0.2)}, anchor=north, legend columns=-1},
      legend to name=sharedlegend
    ]

    \nextgroupplot[ylabel=\gls{mape}, y unit=\si{\%}]
    \addplot table[x=Offset,y=MAPE,col sep=comma] {data/usd/svm/test_mape.csv};
    \addplot table[x=Offset,y=MAPE,col sep=comma] {data/usd/svm/train_mape.csv};

    \nextgroupplot[ylabel=\gls{ds},  y unit=\si{\%}]
    \addplot table[x=Offset,y=Hitrate,col sep=comma] {data/usd/svm/test_hitrate.csv};
    \addplot table[x=Offset,y=Hitrate,col sep=comma] {data/usd/svm/train_hitrate.csv};

    \addlegendentry{Test}
    \addlegendentry{Train}

    \end{groupplot}
  \end{tikzpicture}

  \vspace{1em}
  %\ref{sharedlegend}
  \pgfplotslegendfromname{sharedlegend}
  \caption{\gls{usd} \gls{svr}: \gls{mape} and \gls{ds} for different sample windows.}
\end{figure}

\newpage

The values where the sum of \gls{mape} between the training and test sets was the smallest were selected as exhibiting the best performance. In the case of the \gls{mlp} there was a significant drop in the value of the \gls{mape} with a 30 day sample window. For the \gls{rf} the sample window appeared to have little effect on the forecasting accuracy of the model, as such the sample window of 30 days was selected on the basis of the minimal \gls{oob} error observed. For \gls{svr} the smallest \gls{mape} was found to be with a 30 day sample window.

\subsubsection{Grid search}\label{Grid search}

With the sample window size selected in the previous step a grid search with 5-fold cross-validation was conducted for each of the models.

\begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|}
        \hline
        Parameters & Search grid & Optimium values \\
        \hline
        \hline
        hidden\_layer\_sizes & (50), (100), (200) & (100) \\
        \hline
        activation & tanh & tanh \\
        \hline
        solver & adam, sgd & adam \\
        \hline
        alpha & 0.0001, 0.001, 0.01, 0.1 & 0.01 \\
        \hline
        learning\_rate\_init & 0.001, 0.01 & 0.001 \\
        \hline
        max\_iter & 1000 & 1000 \\
        \hline
    \end{tabular}
    \caption{\gls{usd} \gls{mlp} parameter search grid and found optimum values.}
\end{table}

For the internal cross-validation the negative \gls{mape} was equal to -2.058. For the test set the results were:
\begin{itemize}
    \item \Gls{mape}: 2.960
    \item \Gls{ds}: 0.83
\end{itemize}

\begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|}
        \hline
        Parameters & Search grid & Optimium values \\
        \hline
        \hline
        n\_estimators & 100, 200 & 100 \\
        \hline
        max\_features & 1.0, sqrt, log2, 0.3 & sqrt \\
        \hline
        max\_depth & 1, 3, 7, 10 & 1 \\
        \hline
        min\_samples\_split & 2, 5, 10 & 2 \\
        \hline
        min\_samples\_leaf & 1, 2, 4 & 2 \\
        \hline
    \end{tabular}
    \caption{\gls{usd} \gls{rf} parameter search grid and found optimum values.}
\end{table}

For the internal cross-validation the negative \gls{mape} was equal to -2.264. For the test set the results were:
\begin{itemize}
    \item \Gls{mape}: 2.521
    \item \Gls{ds}: 0.83
\end{itemize}

\begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|}
        \hline
        Parameters & Search grid & Optimium values \\
        \hline
        \hline
        tol & 0.00001, 0.0001, 0.001, 0.01, 0.1 & 0.1 \\
        \hline
        C & 1, 10, 100 & 100 \\
        \hline
        epsilon & 0.1, 0.01, 0.001 & 0.1 \\
        \hline
        gamma & scale, auto & auto \\
        \hline
        kernel & rbf, poly, sigmoid & poly \\
        \hline
    \end{tabular}
    \caption{\gls{usd} \gls{svr} parameter search grid and found optimum values.}
\end{table}

For the internal cross-validation the negative \gls{mape} was equal to -1.398. For the test set the results were:
\begin{itemize}
    \item \Gls{mape}: 1.524
    \item \Gls{ds}: 0.83
\end{itemize}

